---
title: "R Notebook"
output: html_notebook
---

# The Dataset

The following piece of script needs the precalculated datasets placed in the same directory.The dataset is downloadable from the file sharing service detailed in the README.md file or recreated following the steps there.

The script below does a bunch of calculations on all the input files:
  * Creates lead quantities, so we know the outcome and behaviour of the next bet
  * Create user aggregates, like balance, total sum ammount, total win ammount
  * Calculates absolute and relative changes of game descriptiors between consequent games
  * Labels and calculates game streaks (winning/losing) length and extent
  * Prepares scaled version of the prepared quantities (mostly only centered)
  
Important quantities created (for this article that is):
  * time_bet - unixtime of the bet transaction (timestamp of the block)
  * bet_value - transformed from bet ammount in satoshi to bitcoin (1 satoshi = 1e-8 BTC)
  * streak - length of streak (win/loss) 0 signs that the outcome is different from the previous game's
  * label - identifier of the sample, check the bitcoin_pricehistory_lowvar.R script for details
  * pwin - implied probability of winning (0-100), set from the scraped data, defined using the address the bet has been sent to
  * status_pred - status of the next game played 
  * pwin_pred - pwin of the following game
  * game_changed - logical variable, TRUE if a different game (pwin/odds/mulitplier) have been bet on after the current game

```{r echo=FALSE, cache=FALSE, results=FALSE, warning=FALSE, comment=FALSE, warning=FALSE}
# require(groundhog)
# groundhog.day="2021-03-01"
# pkgs=c('tidyverse','ggpubr')
# groundhog.library(pkgs, groundhog.day)
library(tidyverse)

#change this to wherever you put your stuff
#WORKDIR <- "/Users/sandorm/bce-bitluck" #set this for your own working directory
WORKDIR <- "C:/Users/sampa/Documents/bce-bitluck" #set this for your own working directory

normalizeToBitcoin <- function(user_bets){
  user_bets %>% 
    mutate(bet_value = bet_value * 1e-8,
           answer_value = answer_value * 1e-8)
}

aggregateUser <- function(user_bets){
  user_bets %>% 
    mutate(last_outcome = answer_value - bet_value,
           win_answer_value = ifelse(status == "win",answer_value - bet_value,0),
           loose_answer_value = ifelse(status == "loss",answer_value - bet_value,0)) %>%
    mutate(dice_balance = cumsum(answer_value - bet_value),
           winsum = cumsum(win_answer_value),
           loosesum = cumsum(loose_answer_value))
  
}

addlag <- function(user_bets){
  user_bets %>% 
    mutate(bet_pred = lead(bet_value, n = 1),
           game_pred = lead(odds, n = 1),
           time_nextbet = lead(time_bet, n = 1),
           status_pred = lead(status, n = 1))
}
# gametable mined from https://web.archive.org/web/20130403062213/http://www.satoshidice.com/
pwin_to_prob <- c(
  "lessthan","64000","1dice9wVtrKZTBbAZqz1XiTmboYyvpD3t","97.6563","1.004","1.900","98.100","0.0010","250.0000",
  "lessthan","60000","1diceDCd27Cc22HV3qPNZKwGnZ8QwhLTc","91.5527","1.071","1.900","98.100","0.0010","250.0000",
  "lessthan","56000","1dicegEArYHgbwQZhvr5G9Ah2s7SFuW1y","85.4492","1.147","1.900","98.100","0.0010","250.0000",
  "lessthan","52000","1dicec9k7KpmQaA8Uc8aCCxfWnwEWzpXE","79.3457","1.235","1.900","98.100","0.0010","250.0000",
  "lessthan","48000","1dice9wcMu5hLF4g81u8nioL5mmSHTApw","73.2422","1.338","1.900","98.100","0.0010","250.0000",
  "lessthan","32768","1dice97ECuByXAvqXpaYzSaQuPVvrtmz6","50.0000","1.957","1.900","98.100","0.0010","250.0000",
  "lessthan","32000","1dice8EMZmqKvrGE4Qc9bUFf9PX3xaYDp","48.8281","2.004","1.900","98.100","0.0010","250.0000",
  "lessthan","24000","1dice7W2AicHosf5EL3GFDUVga7TgtPFn","36.6211","2.670","1.900","98.100","0.0010","250.0000",
  "lessthan","16000","1dice7fUkz5h4z2wPc1wLMPWgB5mDwKDx","24.4141","4.003","1.900","98.100","0.0010","219.6353",
  "lessthan","12000","1dice7EYzJag7SxkdKXLr8Jn14WUb3Cf1","18.3105","5.335","1.900","98.100","0.0010","152.1241",
  "lessthan","8000","1dice6YgEVBf88erBFra9BHf6ZMoyvG88","12.2070","8.000","1.900","98.100","0.0010","120.9568",
  "lessthan","6000","1dice6wBxymYi3t94heUAG6MpG5eceLG1","9.1553","10.666","1.900","98.100","0.0010","120.9090",
  "lessthan","4000","1dice6GV5Rz2iaifPvX7RMjfhaNPC8SXH","6.1035","15.996","1.900","98.100","0.0010","77.9318",
  "lessthan","3000","1dice6gJgPDYz8PLQyJb8cgPBnmWqCSuF","4.5776","21.326","1.900","98.100","0.0010","57.4951",
  "lessthan","2000","1dice6DPtUMBpWgv8i4pG8HMjXv9qDJWN","3.0518","31.987","1.900","98.100","0.0010","59.3926",
  "lessthan","1500","1dice61SNWEKWdA8LN6G44ewsiQfuCvge","2.2888","42.647","1.900","98.100","0.0010","44.1897",
  "lessthan","1000","1dice5wwEZT2u6ESAdUGG6MHgCpbQqZiy","1.5259","63.968","1.900","98.100","0.0010","29.2270",
  "lessthan","512","1dice4J1mFEvVuFqD14HzdViHFGi9h4Pp","0.7813","124.933","1.900","98.100","0.0010","14.8497",
  "lessthan","256","1dice3jkpTvevsohA4Np1yP4uKzG1SRLv","0.3906","249.861","1.900","98.100","0.0010","7.3951",
  "lessthan","128","1dice37EemX64oHssTreXEFT3DXtZxVXK","0.1953","499.717","1.900","98.100","0.0010","3.6902",
  "lessthan","64","1dice2zdoxQHpGRNaAWiqbK82FQhr4fb5","0.0977","999.429","1.900","98.100","0.0010","3.2050",
  "lessthan","32","1dice2xkjAAiphomEJA5NoowpuJ18HT1s","0.0488","1998.853","1.900","98.100","0.0010","1.6017",
  "lessthan","16","1dice2WmRTLf1dEk4HH3Xs8LDuXzaHEQU","0.0244","3997.701","1.900","98.100","0.0010","0.8006",
  "lessthan","8","1dice2vQoUkQwDMbfDACM1xz6svEXdhYb","0.0122","7995.397","1.900","98.100","0.0010","0.4002",
  "lessthan","4","1dice2pxmRZrtqBVzixvWnxsMa7wN2GCK","0.0061","15990.789","1.900","98.100","0.0010","0.2001",
  "lessthan","2","1dice1Qf4Br5EYjj9rnHWqgMVYnQWehYG","0.0031","31981.573","1.900","98.100","0.0010","0.1000",
  "lessthan","1","1dice1e6pdhLzzWQq7yMidf6j8eAg7pkY","0.0015","64000.000","1.844","98.156","0.0010","0.0500") %>%
  matrix(ncol = 9, byrow = T, dimnames = list(NULL,c("Name","Name2","Address","pwin","odds","House cut (%)","Expected return (%)","Min bet (BTC)","Max bet (BTC)"))) %>%
  as_tibble() %>%
  select(odds,pwin) %>%
  type_convert()

extractAndPrepareUser <- function(x,...){
  ret <- x %>%
    arrange(time_bet) %>% 
    normalizeToBitcoin() %>% 
    addlag() %>% 
    aggregateUser() %>% 
    left_join(pwin_to_prob, by = "odds") %>%
    left_join(pwin_to_prob %>% rename(pwin_pred = pwin), by = c("game_pred" = "odds")) %>%
    mutate(bet_changed = bet_pred != bet_value,
           game_changed = game_pred != odds,
           exp_changed = bet_pred*game_pred != bet_value*odds,
           bet_changed_factor = sign(bet_value-bet_pred) %>% factor(levels = c(0,1,-1)),
           game_changed_factor = sign(game_pred-odds) %>% factor(levels = c(0,1,-1)),
           exp_changed_factor = sign(bet_pred*game_pred - bet_value*odds) %>% factor(levels = c(0,1,-1)),
           expectation = bet_pred*game_pred,
           exp_change_ratio = log(bet_pred*game_pred/(bet_value*odds)),
           bet_change_ratio = log(bet_pred/bet_value),
           game_diff_ratio = log(pwin_pred/pwin),
           win_cumm_log = log10(1+abs(winsum)),
           loose_cumm_log = log10(1+abs(loosesum)),
           dice_balance_log = log10(1+abs(dice_balance)),
           dice_balance_sign = as.character(sign(dice_balance)),
           streak = 0,
           difftime = time_nextbet - time_bet,
           stayed = !((time_nextbet - time_bet) > 4*3600) & (timediff < 3600),
           stay_length = 1,
           game_streak = 0,
           streak_total_length = 0,
           streak_balance = 0) %>%
    drop_na()

  
  for(i in 1:nrow(ret)){
    if(i == 1){
      ret$streak[i] = 0
      ret$stay_length[i] = 1
      ret$game_streak[i] = 0
      ret$streak_balance[i] = 0
    }else{ 
      #win streak
      if(ret$status[i] == ret$status[i-1]){
        ret$streak[i] = ret$streak[i-1]+1
        ret$streak_balance[i] = ret$streak_balance[i-1]+ret$answer_value[i]-ret$bet_value[i]
      }else{
        ret$streak[i] = 0
        ret$streak_balance[i] = 0
      }
      #stay length 
      if(ret$stayed[i]){
        ret$stay_length[i] = ret$stay_length[i-1]+1
      }else{
        ret$stay_length[i] = 1
      }
      #game streak 
      if(!ret$game_changed[i]){
        ret$game_streak[i] = ret$game_streak[i-1]+1
      }else{
        ret$game_streak[i] = 0
      }
    }
  }
  
  for(i in nrow(ret):1){
    if(i == nrow(ret)){
      ret$streak_total_length[i] = ret$streak[i]
      streak_total_length = ret$streak[i]
    }else{
      if(streak_total_length & ret$streak[i]){
        ret$streak_total_length[i] = streak_total_length
      }else if(streak_total_length & !ret$streak[i]){
        ret$streak_total_length[i] = streak_total_length
        streak_total_length = 0
      }else if(!streak_total_length & ret$streak[i]){
        streak_total_length = ret$streak[i]
        ret$streak_total_length[i] = streak_total_length
      }else{
        ret$streak_total_length[i] = streak_total_length
      }
    }
  }
  return(ret %>% arrange(time_bet))
}

prepare_user_table <- function(user_bets_matched, min_games = 2, ...){
  user_bets_matched %>% 
    drop_na() %>%
    group_by(userID) %>% 
    mutate(ngamez = n()) %>% 
    filter(ngamez > min_games) %>%
    group_modify(extractAndPrepareUser) %>%
    ungroup()
}

#loading and processing data
#datasets should be in calcdata folder under working directory
setwd(WORKDIR)
# user_bets_matched_list <- tibble(
#   label = c("A","B","C","D","E"),
#   start_date = c("2012-05-02", "2012-09-17", "2012-12-17", "2013-05-04", "2013-09-11"),
#   end_date = c("2012-05-22", "2012-10-07", "2013-01-06", "2013-05-24", "2013-10-01")) %>%
#   mutate(user_bets_matched = map2(start_date, end_date, function(x,y)read.csv(file = paste0("calcdata/satoshidice_bets_matched__",x,"_to_",y,"_clean.csv")))) %>%
#   mutate(users_table = map(user_bets_matched, prepare_user_table))
##
## Loading a single dataset a time makes runtime friendlier
##
user_bets_matched <- read.csv(file = "calcdata/satoshidice_bets_matched__2013-09-11_to_2013-10-01_clean.csv") %>%
  prepare_user_table()



```

# Creating split datasets for the trainig 

```{r}
#2013-05-24 is the end of the sample so if we want to predict 7 days after a 3 day period where the 1 day had play which meens the first playdate has to be before 2013-05-16

user_play_datemap <- user_bets_matched %>%
  mutate(date = time_bet %>% 
           as.POSIXct(origin = "1970-01-01") %>% 
           as.Date()) %>%
  group_by(userID,date) %>%
  summarize(nplays = n()) %>%
  arrange(userID,date) %>%
  mutate(nthday = 1:n()) %>%
  ungroup(date) %>%
  mutate(first_betday = min(date)) %>%
  ungroup() %>%
  filter(first_betday < max(date)-as.difftime(10, unit="days"))

user_play_datemap %>% pull(userID) %>% unique() %>% length()

user_bets_matched %>%
  mutate(date = time_bet %>% 
           as.POSIXct(origin = "1970-01-01") %>% 
           as.Date())  %>% 
  mutate(log10_bet_value = log10(bet_value),
         payout_value = answer_value - bet_value) %>%
  group_by(userID, date) %>%
  summarize(ngames = log10(n()),
            session = (sum(difftime > 3600)+1),
            pwin_median = median(pwin),
            pwin_width = max(pwin)-min(pwin),
            bet_mean_log10 = mean(log10_bet_value),
            bet_max_log10 = max(log10_bet_value),
            payout_total = sum(payout_value)) %>%
  ungroup() %>%
  select(ngames, session, pwin_median, pwin_width, bet_max_log10, bet_mean_log10, payout_total) %>%
  GGally::ggpairs(progress = F)

n_obs_days <- 3
n_predict_days <- 7

userdays_observe <- user_play_datemap %>%
  mutate(observe = ifelse(nthday < n_obs_days+1,T,F)) %>%
  mutate(predict = ifelse(nthday %>% between(n_obs_days+1 ,n_obs_days+n_predict_days),T,F)) %>%
  select(userID, date, observe, predict)
  

user_observation_set <- user_bets_matched %>%
  mutate(date = time_bet %>% 
           as.POSIXct(origin = "1970-01-01") %>% 
           as.Date()) %>%
  left_join(userdays_observe %>% select(-predict), by = c("userID","date")) %>%
  filter(observe)

user_predict_set <- user_bets_matched %>%
  mutate(date = time_bet %>% 
           as.POSIXct(origin = "1970-01-01") %>% 
           as.Date()) %>%
  left_join(userdays_observe %>% select(-observe), by = c("userID","date")) %>%
  filter(predict)

obs_users <- user_observation_set %>% pull(userID) %>% unique() 
pred_users <- user_predict_set %>% pull(userID) %>% unique() 

```

## Creating group labels for the predict part

```{r}
#prepare dataset for clustering

user_predict_stats <- user_predict_set %>%
  mutate(log10_bet_value = log10(bet_value),
         payout_value = answer_value - bet_value) %>%
  group_by(userID) %>%
  summarize(ngames = log10(n()),
            ndays = n_distinct(date),
            session_per_day = (sum(difftime > 3600)+1)/n_distinct(date),
            pwin_median = median(pwin),
            pwin_width = max(pwin)-min(pwin),
            bet_mean_log10 = mean(log10_bet_value),
            bet_max_log10 = max(log10_bet_value),
            payout_total = sum(payout_value),
            payout_total_to_median_bet = sum(payout_value)/median(bet_value),
            .groups = "drop")

#run trim kmeans

scaled_predict_data <- user_predict_stats %>%
  select(ngames, ndays, session_per_day, pwin_median, pwin_width, bet_max_log10, bet_mean_log10, payout_total) %>%
  scale()

set.seed(123)
km.pred <- trimcluster::trimkmeans(scaled_predict_data, 2, trim = .01, runs = 100)

km.pred$classification %>% table()

#check clustering descriptors

scaled_predict_data_2 <- scaled_predict_data %>% 
  bind_cols(cluster = km.pred$classification) %>%
  filter(cluster != 3) %>%
  select(-cluster)

classific <- km.pred$classification[km.pred$classification != 3]

fpc::cluster.stats(d = dist(scaled_predict_data_2), classific)

clValid::dunn(clusters = km.pred$classification, Data = scaled_predict_data, method = "euclidean")

#examine the cluster statistics

predict_data_clustered <- user_predict_stats %>% 
  cbind(cluster = km.pred$classification)

aggregate(predict_data_clustered, by=list(cluster=km.pred$classification), median)

aggregate(predict_data_clustered, by=list(cluster=km.pred$classification), IQR)

predict_data_clustered %>%
  ggplot(aes(x=as.factor(cluster), y=ndays, color=as.factor(cluster))) +
  geom_boxplot()

predict_data_clustered %>%
  ggplot(aes(x=as.factor(cluster), y=10^ngames, color=as.factor(cluster))) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0,500))

predict_data_clustered %>%
  ggplot(aes(x=as.factor(cluster), y=pwin_median, color=as.factor(cluster))) +
  geom_boxplot()

predict_data_clustered %>%
  ggplot(aes(x=as.factor(cluster), y=pwin_width, color=as.factor(cluster))) +
  geom_boxplot()

predict_data_clustered %>%
  ggplot(aes(x=as.factor(cluster), y=payout_total, color=as.factor(cluster))) +
  geom_boxplot() +
  coord_cartesian(ylim = c(-10,2))

predict_data_clustered %>%
  ggplot(aes(x=as.factor(cluster), y=payout_total_to_median_bet, color=as.factor(cluster))) +
  geom_boxplot() +
  coord_cartesian(ylim = c(-100,100))

#KW tests for the target variable over the two clusters (excluding the trim)
predict_data_clustered %>% filter(cluster != 3) %>% {kruskal.test(ngames ~ cluster, data = .)}
predict_data_clustered %>% filter(cluster != 3) %>% {kruskal.test(ndays ~ cluster, data = .)}
predict_data_clustered %>% filter(cluster != 3) %>% {kruskal.test(session_per_day ~ cluster, data = .)}
predict_data_clustered %>% filter(cluster != 3) %>% {kruskal.test(pwin_median ~ cluster, data = .)}
predict_data_clustered %>% filter(cluster != 3) %>% {kruskal.test(pwin_width ~ cluster, data = .)}
predict_data_clustered %>% filter(cluster != 3) %>% {kruskal.test(bet_max_log10 ~ cluster, data = .)}
predict_data_clustered %>% filter(cluster != 3) %>% {kruskal.test(bet_mean_log10 ~ cluster, data = .)}
predict_data_clustered %>% filter(cluster != 3) %>% {kruskal.test(payout_total ~ cluster, data = .)}
```
## Prepare input set for the autoML algorithm

```{r}

#add labels

addict_clusters <- c(1) #this number may change depending on labeling and k

predict_labels <- predict_data_clustered %>%
  select(userID, cluster) %>%
  rbind(tibble(userID = setdiff(obs_users,pred_users),
               cluster = 0
               )
        ) %>%
  mutate(label_stay = ifelse(cluster != 0, T, F),
         label_addict = ifelse(cluster %in% addict_clusters, T, F))

#initiating h2o server
library(h2o)
h2o.init()


#preapre observation set
user_observation_set_daily_h2o <- user_observation_set %>%
  mutate(log10_bet_value = log10(bet_value),
         payout_value = answer_value - bet_value) %>%
  group_by(userID, date) %>%
  summarize(ngames = log10(n()),
            session = (sum(difftime > 3600)+1),
            pwin_median = median(pwin),
            pwin_width = max(pwin)-min(pwin),
            bet_mean_log10 = mean(log10_bet_value),
            bet_max_log10 = max(log10_bet_value),
            payout_total = sum(payout_value)) %>%
  mutate(nthday = 1:n()) %>%
  mutate(delta_ngames = ngames-lag(ngames),
         delta_bet_mean_log10 = bet_mean_log10-lag(bet_mean_log10)) %>%
  ungroup() %>%
  select(-date) %>%
  pivot_wider(
    names_from = nthday,
    names_sep = "_",
    values_from = c(ngames, session, pwin_median, pwin_width, bet_max_log10, bet_mean_log10, payout_total, delta_ngames, delta_bet_mean_log10),
    values_fill = list(ngames = 0, 
                       session = 0, 
                       pwin_median = 50, 
                       pwin_width = 0, 
                       bet_max_log10 = -5, 
                       bet_mean_log10 = -5, 
                       payout_total = 0,
                       delta_ngames = NA,
                       delta_bet_mean_log10 = NA)
  ) %>%
  left_join(predict_labels, by = "userID") %>%
  mutate(cluster = as.factor(cluster)) %>%
  select(-delta_ngames_1) %>%
  select(-delta_bet_mean_log10_1) %>%
  as.h2o()

#check what we have made
h2o.no_progress() 
h2o.describe(user_observation_set_daily_h2o)
```
## Prepare and run autoML for retention
```{r}
y <- "label_stay"
x <- setdiff(names(user_observation_set_daily_h2o), c("userID","cluster", "label_stay","label_addict","label_addict_smallbet","label_addict_bigbet"))

aml <- h2o.automl(y = y, x = x,
                  training_frame = user_observation_set_daily_h2o,
                  max_models = 10,
                  seed = 123,
                  max_runtime_secs = 300)
```

## Explore the generated model for retention
```{r}
aml@leaderboard
aml@leaderboard$model_id[1] %>% as.data.frame() %>% pull() %>% h2o.getModel()
aml@leaderboard$model_id[1] %>% as.data.frame() %>% pull() %>% h2o.getModel() %>% h2o.varimp_plot() #this will only work for non-ensemble top models
```
## Prepare and run autoML for problem group
```{r}
y <- "label_addict"
x <- setdiff(names(user_observation_set_daily_h2o), c("userID","cluster", "label_stay","label_addict","label_addict_smallbet","label_addict_bigbet"))

aml <- h2o.automl(y = y, x = x,
                  training_frame = user_observation_set_daily_h2o,
                  max_models = 10,
                  seed = 123,
                  max_runtime_secs = 300)
```

## Explore the generated model for problem group
```{r}
aml@leaderboard
aml@leaderboard$model_id[1] %>% as.data.frame() %>% pull() %>% h2o.getModel()
aml@leaderboard$model_id[1] %>% as.data.frame() %>% pull() %>% h2o.getModel() %>% h2o.varimp_plot() #this will only work for non-ensemble top models
```
